The Concurrent Calculi Formalisation Benchmark

### 2024-02-20 short-paper/short-paper.tex
first draft of scope extrusion -- needs some more work, but not today

\subsubsection{Linearity.}

@@

For our challenge on linear reasoning we have chosen a type safety
theorem for a system using session types, combining as usual subject reduction with absence of errors.  This allows us to introduce
linear reasoning with few definitions.  With this focus, we forgo
channel delegation: input and output receives/sends only values, which
do not include channels, hence the topology of the communication
network described by a process cannot change.  We do not allow
recursion or replication in the syntax, hence no infinite behaviours
can be expressed. In particular, we only model linear (as opposed to
shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names, highlighting their
duality within the type system. The latter enforces
 two invariants, namely that no endpoint is used simultaneously by parallel processes and
 that the two endpoints of the same session have dual types.
The operational semantics is formulated in term of reductions rather
than via a transition system, allowing implementors to explore
alternative ways to encode structural congruence\footnote{\alb{Should we mention this in challenge 2?}}.

The statement of subject reduction is the usual one, encompassing type
preservation for structural congruence. W.r.t.\ type safety, we have
chosen a simple notion of what an error is: well-typed processes are a
well-formed in the sense that they do not use endpoints in a non-dual
way.  While more sophisticated notions of well-formedness that
prevent \eg deadlocks are interesting, the proofs of those properties
would distract from the core linear aspects of the challenge.

@@

\subsubsection{Scope extrusion.}
\alb{This needs some work: we do not describe the challenge at all and the citations are not the one I had in mind}

The encoding of binding constructs has been extensively studied
w.r.t.\ the $\lambda$-calculi, but process calculi present additional
challenges.  $\pi$-calculi typically include several different binding
constructs: inputs bind a received name or value, recursive processes
bind recursion variables, and restrictions bind names.  The first two
act similarly to the binders known from $\lambda$-calculi, but
restrictions may be more challenging due to scope extrusion.

\cite{Maksimovic2015} does not have name restriction, thus dodging the challenge of scope extrusion.
\cite{Castro-Perez2021,Castro2020} note that popular presentations of session type systems are impossible to directly encode using intrinsically \(\alpha\)-convertible terms.
\cite{Castro2020} suggests using several binder constructs to solve this, but notes that this requires proving many variants of the same theorems.

\alb{We need more details about what the challenge is}
This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders in restrictions and input.

The case study \cite{AmbalLS21} compares four approaches to encoding binders in Coq as a first step towards developing tools for working with higher-order process calculi.
They found that working directly with de Bruijn indices was easiest and shortest since the existing approaches developed for $\lambda$-calculus binders do not work well with scope extrusion.
Specifically, the locally nameless representation~\cite{Chargueraud2012} cannot avoid direct manipulation of de Bruijn indices when defining the semantics of scope extrusion; cofinite quantification provides no benefits when working under binders; nominal approaches~\cite{Pitts2003} require explicitly giving and validating sets of free names during scope extrusion; and higher-order abstract syntax (HOAS)~\cite{Pfenning1988} requires additional axioms to work with contexts.

Additionally, many proof assistants lack support for the proof principles required to easily work with scope extrusion, \ie, induction principles on functions to work with HOAS encodings.
Some proof assistants such as Abella  support HOAS natively, but working with open terms as required to encode scope extrusion in these systems is generally very challenging.

\subsubsection{Coinduction.}
Coinduction is fundamental to establish properties of processes that
may never terminate, which is a typical situation.
Common uses for coinduction are to define strong and weak versions of
bisimulation, barbed notions of bisimulation, and prove trace equivalence.
On the other hand, coinduction in proof assistants
is notoriously difficult and its representation
is non-uniform across developments.
E.g., \cite{Castro-Perez2021} notes that the most challenging
part of their mechanisation of multiparty session types
was the convoluted definition of infinite behaviours;
\cite{Pohjola2022} points out that the main difficulty
of the proofs concerning bisimulation lies
in dealing with existential quantifiers; and
\cite{Bengtson2016} wonders whether parametrised
coinduction as developed in Coq \cite{Hur2013} would
have helped their Isabelle mechanisation.

>>>>

\subsection{Linearity}

@@

For our challenge on linear reasoning we have chosen a type safety
theorem for a system using session types, combining as usual subject reduction with absence of errors.  This allows us to introduce
linear reasoning with few definitions.  With this focus, we forgo
channel delegation: input and output receives/sends only values, which
do not include channels, hence the topology of the communication
network described by a process cannot change.  We do not allow
recursion or replication in the syntax, hence no infinite behaviours
can be expressed. In particular, we only model linear (as opposed to
shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names, highlighting their
duality within the type system. The latter enforces
 two invariants, namely that no endpoint is used simultaneously by parallel processes and
 that the two endpoints of the same session have dual types.
The operational semantics is formulated as a standard reduction relation to factor the subject reduction proof,  encompassing type
preservation for structural congruence. W.r.t.\ type safety, we have
chosen a simple notion of what an error is: well-typed processes are a
well-formed in the sense that they do not use endpoints in a non-dual
way.  While more sophisticated notions of well-formedness that
prevent \eg deadlocks are interesting, the proofs of those properties
would distract from the core linear aspects of the challenge.

@@

\subsection{Scope extrusion}

This challenge revolves around the formalization explicit scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be ``extruded'' (\ie,
expanded) to include the receiving process.  
The setting for this challenge is a ``classic'' untyped
\( \pi \)-calculus, where (unlike the calculi in the other challenges)
names can be sent and received, and bound by input constructs.  We define two
different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning
about scope extrusion via structural congruence, allowing implementors to explore
alternative ways to encode the latter (\eg via context or via congruent refinements)
\item A (early) labelled transition system
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.

This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders, whose encoding has been
extensively studied w.r.t.\ the $\lambda$-calculi. Process calculi
present additional challenges: $\pi$-calculi typically include several
different binding constructs: inputs bind a received name or value,
recursive processes bind recursion variables, and restrictions bind
names.  The first two act similarly to the binders known from
$\lambda$-calculi, but restrictions may be more challenging due to
scope extrusion.

Given those peculiarities, the syntax and semantics has been
mechanized from a an early age (see~\cite{Melham94}) with many proof
assistants and so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
(Melham's op.cit.) to basic de Bruijn (cite) and locally nameless
representation~\cite{Chargueraud2012} (\eg~\cite{Castro2020}).
Nominal approaches are also common, see~\cite{Bengtson2009}, but they
may be problematic w.r.t.~proof assistants based on dependent types.

\todo{Sum up some of those issues}
% \cite{Maksimovic2015} does not have
% name restriction, thus dodging the challenge of scope extrusion.
\begin{quotation}
\cite{Castro-Perez2021,Castro2020} note that popular presentations of
session type systems are impossible to directly encode using
intrinsically \(\alpha\)-convertible terms.  \cite{Castro2020}
suggests using several binder constructs to solve this, but notes that
this requires proving many variants of the same theorems.
The case study \cite{AmbalLS21} compares four approaches to encoding
binders in Coq as a first step towards developing tools for working
with higher-order process calculi.  They found that working directly
with de Bruijn indices was easiest and shortest since the existing
approaches developed for $\lambda$-calculus binders do not work well
with scope extrusion.  Specifically, the locally nameless
representation~\cite{Chargueraud2012} cannot avoid direct manipulation
of de Bruijn indices when defining the semantics of scope extrusion;
cofinite quantification provides no benefits when working under
binders; nominal approaches require explicitly giving
and validating sets of free names during scope extrusion;
\end{quotation}

Higher-order abstract syntax (HOAS) has seen extensive and successful
use in formal reasoning~\cite{Despeyroux2000,Honsell2001,Tiu2010}. Its
weak form aligns well with mainstream inductive proof assistants,
simplifying the encoding of typing systems and operational
semantics. However, when addressing more intricate concepts like
bisimulation, HOAS necessitates extensions. These extensions may take
the form of additional axioms~\cite{Honsell2001} or require specialized proof assistants
like Abella, which introduces a dedicated quantifier for handling
properties related to names~\cite{GacekMN11}.

% Higher-order abstract syntax (HOAS) has been extensively and
% successfully used~\cite{Despeyroux2000,Honsell2001,Tiu2010}, since in
% its weak form it is compatible with mainstream inductive proof
% assistants. While this approach disposes of most of the technicalities
% when encoding typing systems or operational semantics, it requires
% extensions when dealing with notions such as bisimulation, either in the
% form of additional axioms~\cite{Honsell2001}, or of dedicated proofs
% assistants such as Abella, which features a specific
% quantifier~\cite{GacekMN11} to handle properties of names.

\subsection{Coinduction}
Coinduction is fundamental to establish properties of processes that
may never terminate, which is a typical situation.
Common uses for coinduction are to define strong and weak versions of
bisimulation, barbed notions of bisimulation, and prove trace equivalence.
On the other hand, coinduction in proof assistants
is notoriously difficult and its representation
is non-uniform across developments.
E.g., \cite{Castro-Perez2021} notes that the most challenging
part of their mechanisation of multiparty session types
was the convoluted definition of infinite behaviours;
\cite{Pohjola2022} points out that the main difficulty
of the proofs concerning bisimulation lies
in dealing with existential quantifiers; and
\cite{Bengtson2016} wonders whether parametrised
coinduction as developed in Coq \cite{Hur2013} would
have helped their Isabelle mechanisation.
